Confidence Intervals for Prevalence Estimates from Complex Surveys with Imperfect Assays

### 2021-09-14 PrevAdjSurvey.tex
Updates from Overleaf

\documentclass[AMA,STIX1COL]{WileyNJD-v2}
\usepackage{moreverb}
% just for todo notes
\setlength {\marginparwidth }{2cm}
\usepackage[colorinlistoftodos]{todonotes}

@@

\title{A demonstration of the \LaTeX\ class file for Wiley NJD Journals\protect\thanks{This is an example for title footnote.}}

@@

\section{Methods}
\todo[inline]{I’m struggling a bit with the overall message of the paper.

In Part 1, we advocate for the Kalish Appendix aka the Korn-Graubard + CD method over the Lang-Reiczigel method in the case of a simple random sample / unweighted survey. This is sort of a special case of the / Korn-Graubard + CD method with k = 1 and the binomial variance assumption, rather than the Poisson variance assumption.

In Part 2, we advocate for Fay-Feuer over Korn-Graubard in the case where
 }
Suppose we have a survey with  \( k \) sampling units, with \( N_1, N_2, \ldots, N_k \) individuals in the population of unit \( i \).
We sample \( n_1, n_2, \ldots, n_k \) individuals via a simple random sample from each population to have an assay performed to determine who has a disease.
Let \( \theta_i \) be the population frequency of positive results for assays performed on individuals from sampling unit \( i \) and \( X_i \) be the number of positive results from an assay performed on \( n_i \) individuals from sampling unit \( i \).
Also let \( \theta_i^* \) be the population frequency of cases in sampling unit \( i \) and \( X_i^* \) be the true number of people with the disease among the \( n_i \) individuals from sampling unit \( i \).
Then \( X_i^* \sim \operatorname{Binomial}(n_i, \theta_i^*) \)
In the case of a perfect assay, \( \theta_i = \theta_i^* \) and \( X_i = X_i^* \).

@@

\begin{equation}
    \beta = \frac{\sum_{i=1}^k N_i \theta_i^*}{\sum_{j=1}^k N_j} = \sum_{i=1}^k w_i \theta_i^*,
\end{equation}

where \( w_i = N_i / \sum_{j=1}^n N_j \).

Suppose the assay is measured on \( m_n \) individuals known to not have the disease and on \( m_p \) individuals known to have have the disease.
Let \( C_n \) and \( C_p \) be the number who test positive from the respective samples.
Assume that the negative and positive controls act like simple random samples from their respective populations and that \( C_n \sim \operatorname{Binomial}(m_n, \phi_n) \) and \( C_p \sim \operatorname{Binomial}(m_p, \phi_p) \), where \( 1 - \phi_n \) is the specificity of the assay and \( \phi_p \) is the sensitivity of the assay.

Let \( \widehat{AP} = \frac{\sum_{i=1}^k N_i \frac{X_i}{n_i}}{\sum_{j=1}^k N_j} = \sum_{i=1}^k w_i \frac{X_i}{n_i} \), \( \widehat{Sp} = 1 - \frac{C_n}{m_n} \), and \( \widehat{Se} = \frac{C_p}{m_p} \).

The first method in this paper is concerned with estimating \( \beta \) in the case where \( k = 1 \), \( \phi_n > 0 \), \( \phi_p < 1 \), i.e. estimating prevalence from a simple random sample with an imperfect assay.
The second method in this paper is concerned with estimating \( \beta \) in the case where \( k > 1 \), \( \phi_n = 0 \), \( \phi_p = 1 \), i.e. estimating prevalence from a weighted sample with a perfect assay.
The final method in this paper is concerned with estimating \( \beta \) in the case where \( k > 1 \), \( \phi_n > 0 \), \( \phi_p < 1 \), i.e. estimating prevalence from a weighted sample with an imperfect assay.

\subsection{Estimating Prevalence from a Simple Random Sample with an Imperfect Assay}

First, we consider the scenario \( k = 1 \), \( \phi_n > 0 \), \( \phi_p < 1 \) and we are concerned with estimating

\begin{equation}
    \beta = \frac{\sum_{i=1}^k N_i \theta_i^*}{\sum_{j=1}^k N_j} = \theta_1^*
\end{equation}

\begin{itemize}
    \item We propose a modification of the melding method described in Fay 2015.
    \item which uses lower and upper confidence distributions on functions of estimators independent estimators to create a confidence interval for prevalence that accounts for variability in \( \widehat{AP} \), \( \widehat{Se} \), and \( \widehat{Sp} \)
    \item We assume all of these come from Binomial distributions.
\end{itemize}

We use the method is described in the Appendix of \cite{Kali:2021}.
This is a modification of the melding method, which uses lower and upper confidence distributions on functions of estimators independent estimators to create a confidence interval for prevalence that accounts for variability in \( \widehat{AP} \), \( \widehat{Se} \), and \( \widehat{Sp} \) \cite{FayP:2015}.

For each of \( \widehat{AP} \), \( \widehat{Se} \), and \( \widehat{Sp} \), we use the lower and upper confidence distributions associated with the exact binomial confidence interval.
In the case of a binomial random variable with \( x \) successes among \( n \) trials, the lower confidence distribution is \( \operatorname{Beta}(x, n - x + 1) \), and the upper confidence distribution is \( \operatorname{Beta}(x + 1, n - x) \).

Letting \( B_{ap}^L, B_{ap}^U, B_{sp}^L, B_{sp}^U, B_{se}^L, B_{se}^U \) be the lower and upper confidence distribtuions associate with the estimators, and \( q(a, W) \) be the \( a \)th quantile of a random variable \( W \), the 95\% confidence interval for \( \beta \) is

\( \left\{ q \left( 0.025, \frac{B_{ap}^L + B_{sp}^L - 1}{B_{se}^U + B_{sp}^L - 1}  \right),  q \left( 0.975, \frac{B_{ap}^U + B_{sp}^U - 1}{B_{se}^L + B_{sp}^U - 1}  \right) \right\}.\)

This method is compared to the method presented in \citet{Lang:2014}.

\subsection{Estimating Prevalence from a Weighted Sample with a Perfect Assay}

@@

@@

>>>>

\documentclass[AMA,STIX1COL]{WileyNJD-v2}
\usepackage{moreverb}

\DeclareMathOperator{\Var}{Var}

% just for todo notes
\setlength {\marginparwidth }{2cm}
\usepackage[colorinlistoftodos]{todonotes}

@@

\title{Confidence Intervals for Prevalence Estimates from Complex Surveys with Imperfect Assays}

@@

\section{Methods}

Suppose we have a survey with  \( k \) sampling units, with \( N_1, N_2, \ldots, N_k \) individuals in the population of unit \( i \).
We sample \( n_1, n_2, \ldots, n_k \) individuals via a simple random sample from each population to have an assay performed to determine who has a disease.
Let \( X_i \) be the number of positive results from an assay performed on the \( n_i \) individuals from sampling unit \( i \) and assume \( X_i \sim \operatorname{Binomial}(n_i, \theta_i) \), where \( \theta_i \) is the population frequency of positive results for assays performed on individuals from sampling unit \( i \).
Similarly, let \( X_i^* \) be the true number of people with the disease among the \( n_i \) individuals from sampling unit \( i \) and assume \( X_i^* \sim \operatorname{Binomial}(n_i, \theta_i^*) \), where \( \theta_i^* \) is the population frequency of cases in sampling unit \( i \).
In the case of a perfect assay, \( \theta_i = \theta_i^* \).

@@

\begin{equation}
    \beta^* = \frac{\sum_{i=1}^k N_i \theta_i^*}{\sum_{j=1}^k N_j} = \sum_{i=1}^k w_i \theta_i^*,
    \label{eq:pop-prev}
\end{equation}

where \( w_i = N_i / \sum_{j=1}^n N_j \) and, therefore, \( \sum_{i=1}^k w_i = 1 \).

Suppose the assay is measured on \( m_n \) individuals known to not have the disease and on \( m_p \) individuals known to have have the disease.
Let \( C_n \) and \( C_p \) be the number who test positive from the respective samples.
Assume that the negative and positive controls act like simple random samples from their respective populations.
Thus, \( C_n \sim \operatorname{Binomial}(m_n, \phi_n) \) where \( 1 - \phi_n \) is the specificity of the assay, and \( C_p \sim \operatorname{Binomial}(m_p, \phi_p) \), where  \( \phi_p \) is the sensitivity of the assay.

Note the relationship between these parameters:

\begin{align}
\begin{split}
  \beta^*   =&   \sum_{i=1}^k w_i \theta_i^* \\
            =&  \sum_{i=1}^k w_i \left( \frac{\theta_i - \phi_n}{\phi_p - \phi_n} \right) \\
            =&   \frac{\sum_{i=1}^k w_i \theta_i}{\phi_p - \phi_n} - \frac{\phi_n \sum_{i=1}^k w_i}{\phi_p - \phi_n} \\
            =&   \frac{\sum_{i=1}^k w_i \theta_i}{\phi_p - \phi_n} - \frac{\phi_n}{\phi_p - \phi_n}
            \label{eq:long-beta}
\end{split}
\end{align}

Let \( \hat{\theta}_i = \frac{X_i}{n_i} \), \( \hat{\phi}_n = \frac{C_n}{m_n} \), and \( \hat{\phi}_p = \frac{C_p}{m_p} \).
Then a natural estimator for \( \beta^* \) is

\begin{equation}
    \hat{\beta}^* = \frac{\sum_{i=1}^k w_i \hat{\theta}_i}{\hat{\phi}_p - \hat{\phi}_n} - \frac{\hat{\phi}_n}{\hat{\phi}_p - \hat{\phi}_n}.
\end{equation}

This estimator serves as important basis for developing confidence intervals in this work.
Section~\ref{sec:srs-imperfect} is concerned with confidence intervals for  \( \beta^* \) in the case where \( k = 1 \), \( \phi_n > 0 \), \( \phi_p < 1 \), i.e. estimating prevalence from a simple random sample with an imperfect assay.
Section~\ref{sec:weight-perfect} is concerned with confidence intervals for  \( \beta^* \) in the case where \( k > 1 \), \( \phi_n = 0 \), \( \phi_p = 1 \), i.e. estimating prevalence from a weighted sample with a perfect assay.
Section~\ref{sec:weight-imperfect} is concerned with confidence intervals for  \( \beta^* \) in the case where \( k > 1 \), \( \phi_n > 0 \), \( \phi_p < 1 \), i.e. estimating prevalence from a weighted sample with an imperfect assay.

\subsection{Estimating Prevalence from a Simple Random Sample with an Imperfect Assay}
\label{sec:srs-imperfect}

First, we consider the scenario \( k = 1 \), \( \phi_n > 0 \), and \( \phi_p < 1 \).
We develop a confidence interval for the population prevalence, \( \beta^* \).
Because \( k = 1 \), the estimator in Equation~\ref{eq:long-beta} becomes

\begin{equation}
\hat{\beta}^* = \frac{\hat{\theta}_1 - \hat{\phi}_n}{\hat{\phi}_p - \hat{\phi}_n}.
\label{eq:srs-beta-est}
\end{equation}

To create a confidence interval for \( \hat{\beta}^* \), we use a generalization of the melding method \cite{FayP:2015}, which makes use of lower and upper confidence distributions on functions of independent estimators to account for variability in \( \hat{\theta}_1 \), \( \hat{\phi}_n \), and \( \hat{\phi}_p \).

Each estimated component in Equation~\ref{eq:srs-beta-est} is a binomial parameter.
For each of these, we use distributions associated with  the exact binomial confidence interval.
For a binomial experiment with \( x \) successes out of \( n \) trials, the lower confidence distribution \( (B^L) \) is \( \operatorname{Beta}(x, n - x + 1) \), and the upper confidence distribution \( (B^U) \) is \( \operatorname{Beta}(x + 1, n - x)\).

Let \( q(a, W) \) be the \( a \)th quantile of a random variable \( W \).

Then the \( a \)\% confidence interval for \( \beta^* \) is

\begin{equation}
    \left\{ q \left( \frac{1 - a}{2}, \frac{B_{\theta_1}^L + B_{\phi_n}^L }{B_{\phi_p}^U + B_{\phi_n}^L }  \right),  q \left( \frac{1 + a}{2}, \frac{B_{\theta_1}^U + B_{\phi_n}^U}{B_{\phi_p}^L + B_{\phi_n}^U}  \right) \right\}.
\end{equation}

We compare this method to one described in \cite{Lang:2014} as implemented in prevSeSp function in \cite{asht}, which provides approximate confidence intervals for true prevalence when sensitivity and specificity are estimated from independent samples, as they are in the in this section.

This interval is given by
\begin{equation}
\beta_1^{*\prime} + d\beta \pm q\left( \frac{1 + a}{2}, Z \right) \cdot \Var(\beta_1^{*\prime})^{1/2}    
\end{equation}

where
\begin{equation}
   Z \sim N(0,1) 
\end{equation}

\begin{equation}
    d\beta = 2 \cdot q\left( \frac{1 + a}{2}, Z \right)^2 \cdot\left\{ \beta_1^{*\prime} \cdot \frac{\phi_p^\prime (1 - \phi_p^\prime)}{m_p^\prime} - (1 - \beta_1^{*\prime}) \cdot \frac{(1 - \phi_n^\prime) \phi_n^\prime}{m_n^\prime}  \right\}
\end{equation}

\begin{equation}
    \Var(\beta_1^{*\prime}) = \frac{ \frac{\beta_1^{*\prime}(1 - \beta_1^{*\prime})}{n_1} + \left(\beta_1^{*\prime}\right)^2 \frac{\phi_p^\prime (1 - \phi_p^\prime)}{m_p} + \left(1 + \beta_1^{*\prime}\right)^2 \frac{(1 - \phi_n^\prime) \phi_n^\prime}{ m_n}}{(\phi_p^\prime - \phi_n^\prime)^2}
\end{equation}

\begin{equation}
    m_p^\prime = m_p +2
\end{equation}

\begin{equation}
    m_n^\prime = m_n + 2
\end{equation}

\begin{equation}
    \phi_p^\prime = \frac{m_p \cdot \hat{\phi}_p + 1}{m_p + 2}
\end{equation}

\begin{equation}
   1 - \phi_n^\prime = \frac{m_n \cdot (1 - \hat{\phi}_n) + 1}{m_n + 2} 
\end{equation}

\begin{equation}
   \beta_1^{*\prime} = \frac{\beta_1^\prime - \phi_n^\prime}{\phi_p^\prime - \phi_n^\prime} 
\end{equation}

\begin{equation}
    \beta_1^\prime = \frac{n_1 \cdot \hat{\theta}_1 + q\left( \frac{1 + a}{2}, Z \right)^2 / 2}{n_1 + q\left( \frac{1 + a}{2}, Z \right)^2}
\end{equation}

\subsection{Estimating Prevalence from a Weighted Sample with a Perfect Assay}
\label{sec:weight-perfect}
Next, we develop a confidence interval for the population prevalence, \( \beta^* \), in the scenario \( k > 1 \), \( \phi_n = 0 \), \( \phi_p = 1 \).
Our method is a straightforward adaptation of the one presented in \cite{FayF:1997}, which was developed to create confidence intervals for a population rate which is assumed to a weighted sum of Poisson rate parameters.
We note that for sufficiently large sample size \( n \) and small rate \( \lambda \), a \( \operatorname{Poisson}(n\lambda) \) distribution is approximately equal in distribution to a \( \operatorname{Binomial}(n, \lambda) \) distribution.
Under this assumption, we suggest the confidence interval for \( \beta^* \):

\begin{equation}
    \left( q\left( \frac{1 - a}{2}, G_{\beta^*}^L \right), q \left( \frac{1 + a}{2},  G_{\beta^*}^U \right) \right)
\end{equation}

Where

\begin{equation}
    G_{\beta^*}^L \sim \operatorname{Gamma}\left( \frac{y^2}{v}, \frac{v}{y} \right)
\end{equation}

\begin{equation}
    G_{\beta^*}^U \sim \operatorname{Gamma}\left( \frac{y^{*2}}{v^*}, \frac{v^*}{y^*} \right)
\end{equation}

\begin{equation}
    y = \sum_{i=1}^k \frac{w_i}{n_i} x_i
\end{equation}

\begin{equation}
    v = \sum_{i=1}^k \left( \frac{w_i}{n_i}\right)^2 x_i
\end{equation}

\begin{equation}
    y^* = y + \max\left(\frac{w_1}{n_1}, \ldots, \frac{w_k}{n_k} \right)
\end{equation}

\begin{equation}
    v^* = v + \left\{ \max\left(\frac{w_1}{n_1}, \ldots, \frac{w_k}{n_k} \right) \right\}^2
\end{equation}

We compare this confidence interval to two methods presented in \cite{Dean:2015}.

The first of these is an adaptation of \cite{AgrestiCoull} for the survey setting.
The interval for \( \beta^* \) is given by:

\begin{equation}
    \tilde{p} \pm q\left( \frac{1 + a}{2}, Z \right) \sqrt{\tilde{p}(1 - \tilde{p}) / \tilde{n}}
\end{equation}

where

\begin{equation}
   \tilde{x} = \left( \sum_{i=1}^k w_i \hat{\theta}_i \right) n_{\text{eff}} + c 
\end{equation}

\begin{equation}
   \tilde{n} = n_{\text{eff}} + 2c 
\end{equation}

\begin{equation}
    \tilde{p} = \tilde{x} / \tilde{n}
\end{equation}

\begin{equation}
   c = q\left( \frac{1 + a}{2}, Z \right)^2/2 
\end{equation}

\begin{equation}
   n_{\text{eff}} = \frac{\left( \sum_{i=1}^k w_i \hat{\theta}_i \right) \left(1 - \sum_{i=1}^k w_i \hat{\theta}_i \right)}{\sum_{i=1}^k \frac{w_i^2}{n_i}\hat{\theta}_i} 
   \label{eq:neff}
\end{equation}

In the case where \( \sum_{i=1}^k \frac{w_i^2}{n_i}\hat{\theta}_i = 0 \), we instead let \( n_{\text{eff}} = \sum_{i=1}^k n_i \).

We also compare our suggested method to a modification of \cite{Korn:1998} presented in \cite{Dean:2015}.
This interval is given by

\begin{equation}
    \left( q \left( \frac{1 - a}{2}, B^L_{CP} \right), q \left( \frac{1 + a}{2}, B^U_{CP} \right)  \right)
\end{equation}

where

\begin{equation}
    B^L_{CP} \sim \operatorname{Binomial}\left(n_{\text{eff}} \sum_{i=1}^k w_i \hat{\theta}_i, n_{\text{eff}} \left(1 -  \sum_{i=1}^k w_i \hat{\theta}_i\right) + 1\right)
\end{equation}

\begin{equation}
    B^U_{CP} \sim \operatorname{Binomial}\left(\left(n_{\text{eff}} \sum_{i=1}^k w_i \hat{\theta}_i\right) + 1, n_{\text{eff}} \left(1 - \sum_{i=1}^k w_i \hat{\theta}_i \right)\right)
\end{equation}

with \( n_{\text{eff}} \) defined in Equation~\ref{eq:neff}.

@@

\label{sec:weight-imperfect}

@@

THERE IS A SMALL ERROR IN THESE RESULTS. RERUNNING ON BIOWULF NOW. I DON'T EXPECT THE SUBSTANCE TO CHANGE.
